---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  eval = as.logical(Sys.getenv("ODBC_EVAL_README", "false"))
)
```

# odbc
<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/odbc)](https://cran.r-project.org/package=odbc)
[![R-CMD-check](https://github.com/r-dbi/odbc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-dbi/odbc/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/r-dbi/odbc/branch/main/graph/badge.svg)](https://app.codecov.io/gh/r-dbi/odbc?branch=main)
<!-- badges: end -->

The goal of the odbc package is to provide a DBI-compliant interface to [Open
Database
Connectivity](https://learn.microsoft.com/en-us/sql/odbc/microsoft-open-database-connectivity-odbc?view=sql-server-ver15)
(ODBC) drivers. This allows for an efficient, easy to setup connection to any
database with ODBC drivers available, including [SQL
Server](https://www.microsoft.com/en-us/sql-server/),
[Oracle](https://www.oracle.com/database), [MySQL](https://www.mysql.com/),
[PostgreSQL](https://www.postgresql.org/), [SQLite](https://sqlite.org/index.html) and
others. The implementation builds on the
[nanodbc](https://nanodbc.github.io/nanodbc/) C++ library.

## Installation 

Install the latest release of odbc from CRAN with the following code:

```r
install.packages("odbc")
```

To get a bug fix or to use a feature from the development version, you can install the development version of odbc from GitHub:

```r
# install.packages("pak")
pak::pak("r-dbi/odbc")
```

odbc and its dependencies use C++11 features. Therefore, you need [gcc 4.8](https://gcc.gnu.org/), [clang 3.3](https://clang.llvm.org), or [Rtools 3.3](https://CRAN.R-project.org/bin/windows/Rtools/) or later.

## Usage

All of the following examples assume you have already created a connection `con`. See the "Installing and Configuring Drivers" vignette for more information on establishing a connection.

<!-- TODO: link to the deployed vignette above -->

`dbListTables()` is used for listing all existing tables in a database.

```r
dbListTables(con)

# List tables beginning with f
dbListTables(con, table_name = "f%")

# List all fields in the 'flights' database
dbListFields(con, "flights")
```

`dbReadTable()` will read a full table into an R `data.frame()`.

```r
data <- dbReadTable(con, "flights")
```

`dbWriteTable()` will write an R `data.frame()` to an SQL table.

```r
dbWriteTable(con, "iris", iris)
```

`dbGetQuery()` will submit a query and fetch the results. It is also possible
to submit the query and fetch separately with `dbSendQuery()` and `dbFetch()`.
The `n` argument to `dbFetch()` can be used to fetch only the part of a query result (the next *n* rows).

```r
result <- dbSendQuery(con, "SELECT flight, tailnum, origin FROM flights ORDER BY origin")

# Retrieve the first 100 results
first_100 <- dbFetch(result, n = 100)

# Retrieve the rest of the results
rest <- dbFetch(result)
```

## Benchmarks

The odbc package is often much faster than the existing
[RODBC](https://cran.r-project.org/package=RODBC) and DBI compatible
[RODBCDBI](https://cran.r-project.org/package=RODBCDBI) packages.

### Reading

Reading a table from a SQL Server database with the 'flights' dataset (336,776
rows, 19 columns) of the package
[nycflights13](https://github.com/tidyverse/nycflights13).

```{r}
# First using RODBC / RODBCDBI
library(DBI)
rodbc <- dbConnect(RODBCDBI::ODBC(), dsn = "MicrosoftSQLServer", user = Sys.getenv("SQLSERVER_UID"), password = Sys.getenv("SQLSERVER_PWD"))
system.time(rodbc_result <- dbReadTable(rodbc, "flights"))
#>    user  system elapsed 
#>  13.986   1.173  15.192

# Now using odbc
odbc <- dbConnect(odbc::odbc(), dsn = "MicrosoftSQLServer", UID = Sys.getenv("SQLSERVER_UID"), PWD = Sys.getenv("SQLSERVER_PWD"))
system.time(odbc_result <- dbReadTable(odbc, "flights"))
#>    user  system elapsed 
#>   1.334   0.065   1.406

tibble::as_tibble(odbc_result)
#> # A tibble: 336,776 x 19
#>     year month   day dep_time sched_dep_time dep_delay arr_time
#>    <int> <int> <int>    <int>          <int>     <dbl>    <int>
#>  1  2013     1     1      517            515         2      830
#>  2  2013     1     1      533            529         4      850
#>  3  2013     1     1      542            540         2      923
#>  4  2013     1     1      544            545        -1     1004
#>  5  2013     1     1      554            600        -6      812
#>  6  2013     1     1      554            558        -4      740
#>  7  2013     1     1      555            600        -5      913
#>  8  2013     1     1      557            600        -3      709
#>  9  2013     1     1      557            600        -3      838
#> 10  2013     1     1      558            600        -2      753
#> # … with 336,766 more rows, and 12 more variables: sched_arr_time <int>,
#> #   arr_delay <dbl>, carrier <chr>, flight <int>, tailnum <chr>,
#> #   origin <chr>, dest <chr>, air_time <dbl>, distance <dbl>, hour <dbl>,
#> #   minute <dbl>, time_hour <dttm>

identical(dim(rodbc_result), dim(odbc_result))
#> [1] TRUE

rm(rodbc_result, odbc_result, odbc, rodbc)
gc(verbose = FALSE)
```

### Writing

Writing the same dataset to the database.

```{r, echo = FALSE, results = "hide"}
odbc <- dbConnect(odbc::odbc(), dsn = "MicrosoftSQLServer", UID = Sys.getenv("SQLSERVER_UID"), PWD = Sys.getenv("SQLSERVER_PWD"))
if (dbExistsTable(odbc, "flights2")) { dbRemoveTable(odbc, "flights2") }
if (dbExistsTable(odbc, "flights3")) { dbRemoveTable(odbc, "flights3") }
rm(odbc)
gc(verbose = FALSE)
```

```{r, cache = TRUE}
library(nycflights13)
# rodbc does not support writing timestamps natively, so we remove that column
flights2 <- as.data.frame(flights[, names(flights) != "time_hour"])
rodbc <- dbConnect(RODBCDBI::ODBC(), dsn = "MicrosoftSQLServer", user = Sys.getenv("SQLSERVER_UID"), password = Sys.getenv("SQLSERVER_PWD"))
system.time(dbWriteTable(rodbc, "flights2", flights2))
#>    user  system elapsed
#>  11.891   6.269 765.269

# Now using odbc
odbc <- dbConnect(odbc::odbc(), dsn = "MicrosoftSQLServer", UID = Sys.getenv("SQLSERVER_UID"), PWD = Sys.getenv("SQLSERVER_PWD"))
flights3 <- as.data.frame(flights)
system.time(dbWriteTable(odbc, "flights3", flights3))
#>    user  system elapsed
#>   2.617   0.275   9.649
```
