---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# odbconnect
[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)
[![Travis-CI Build Status](https://travis-ci.org/rstats-db/odbconnect.svg?branch=master)](https://travis-ci.org/rstats-db/odbconnect)
[![Coverage Status](https://img.shields.io/codecov/c/github/rstats-db/odbconnect/master.svg)](https://codecov.io/github/rstats-db/odbconnect?branch=master)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/2bnahh7qg5iu7197?svg=true)](https://ci.appveyor.com/project/hadley/odbconnect-cqvmd)

The goal of odbconnect is to provide a DBI-compliant interface to [Open
Database
Connectivity](https://msdn.microsoft.com/en-us/library/ms710252(v=vs.85).aspx)
(ODBC) drivers. This gives a efficient, easy to setup connection to any
database with ODBC drivers available, including [SQL
Server](https://www.microsoft.com/en-us/sql-server/),
[Oracle](https://www.oracle.com/database), [MySQL](https://www.mysql.com/),
[PostgreSQL](https://www.postgresql.org/), [SQLite](https://sqlite.org/) and
others. The implementation builds on the
[nanodbc](http://nanodbc.lexicalunit.com/) C++ library.

-   [Installation](#installation)
    -   [Windows](#windows)
    -   [MacOS](#macos)
    -   [Linux - Debian / Ubuntu](#linux---debian-ubuntu)
    -   [R](#r)
-   [Connecting to a Database](#connecting-to-a-database)
    -   [Connection Strings](#connection-strings)
    -   [DSN Configuration files](#dsn-configuration-files)
-   [Usage](#usage)
    -   [Table and Field information](#table-and-field-information)
    -   [Reading](#reading)
    -   [Writing](#writing)
    -   [Querying](#querying)
-   [Benchmarks](#benchmarks)
    -   [Reading](#reading-1)
    -   [Writing](#writing-1)

## Installation
For Unix and MacOS ODBC drivers should be compiled against
[unixODBC](http://www.unixodbc.org/). Drivers compiled against [iODBC](http://www.iodbc.org/) _may_ also
work, but are not fully supported.

### Windows
Windows is bundled with ODBC libraries however drivers for each database
need to be installed separately. Windows ODBC drivers typically include a
installer that needs to be run and will install the driver to the proper
locations.

### MacOS
[homebrew](http://brew.sh/) can be used to easily install database drivers on MacOS.

#### UnixODBC - Required for all databases
```shell
# Install the unixODBC library
brew install unixodbc
```

#### Common DB drivers
```shell
# SQL Server ODBC Drivers (Free TDS)
brew install freetds --with-unixodbc

# PostgreSQL ODBC ODBC Drivers
brew install psqlodbc

# MySQL ODBC Drivers (and database)
brew install mysql

# SQLite ODBC Drivers
brew install sqliteodbc
```

### Linux - Debian / Ubuntu
[apt-get](https://wiki.debian.org/Apt) can be used to easily install database
drivers on Linux distributions which support it, such as Debian and Ubuntu.

#### UnixODBC - Required for all databases
```shell
# Install the unixODBC library
apt-get install unixodbc unixodbc-dev
```

#### Common DB drivers
```shell
# SQL Server ODBC Drivers (Free TDS)
apt-get install tdsodbc

# PostgreSQL ODBC ODBC Drivers
apt-get install odbc-postgresql

# MySQL ODBC Drivers
apt-get install libmyodbc

# SQLite ODBC Drivers
apt-get install libsqliteodbc
```

### R
odbconnect is not yet available on CRAN, however
[devtools](https://cran.r-project.org/package=devtools) can be used to install
the latest version.
```R
# install.packages(devtools)
devtools::install_github("rstats-db/odbconnect")
```

## Connecting to a Database
Databases can be connect by specifying a connection string directly, or with
DSN Configuration files.

### Connection Strings

Specify a connection string as named arguments directly in the `dbConnect()` method.
```r
con <- dbConnect(odbconnect::odbconnect(),
  Driver = "PostgreSQL Driver",
  Database = "test_db",
  uid = "postgres",
  pwd = "password",
  host = "localhost",
  port = 5432)
```

Alternatively you can pass a complete connection string as the
`.connection_string` argument. [The Connection Strings Reference](www.connectionstrings.com) is a useful resource
that has example connection strings for a large variety of databases.
```r
con <- dbConnect(odbconnect::odbconnect(),
  .connection_string = "Driver={PostgreSQL Driver};Uid=postgres;Pwd=password;Host=localhost;Port=5432;Database=test_db;")
```

### DSN Configuration files

ODBC configuration files are another option to specify connection parameters
and allow one to use a Data Source Name (DSN) to make it easier to connect to a database.

```r
con <- dbConnect(odbconnect::odbconnect(), "PostgreSQL")
```

#### Windows
The [ODBC Data Source
Administrator](https://msdn.microsoft.com/en-us/library/ms714024(v=vs.85).aspx)
application is used to manage ODBC data sources on Windows.

#### MacOS / Linux
On MacOS and Linux there are two separate text files that need to be edited.
UnixODBC includes an command line executable `odbcinst` which can be used to
query and modify the DSN files. However these are plain text files you
can also edit by hand if desired.

There are two different files used to setup the DSN information.
- `odbcinst.ini` - which defines driver options
- `odbc.ini` - which defines connection options

The DSN configuration files can be defined globally for all users of the
system, often at
`/etc/odbc.ini` or `/opt/local/etc/odbc.ini`, the exact location depends on
what option was used when compiling unixODBC. `odbcinst -j` can be used to find
the exact location. Alternatively the `ODBCSYSINI` environment variable can be
used to specify the location of the configuration files. Ex. `ODBCSYSINI=~/ODBC`

A local DSN file can also be used with the files `~/.odbc.ini` and `~/.odbcinst.ini`.

##### odbcinst.ini
Contains driver information, particularly the name of the driver library.
Multiple drivers can be specified in the same file.
```ini
[PostgreSQL]
Driver          = /usr/local/lib/psqlodbcw.so

[SQLite Driver]
Driver          = /usr/local/lib/libsqlite3odbc.dylib
```

##### odbc.ini
Contains connection information, particularly the username, password, database
and host information. The Driver line corresponds to the driver defined in
`odbcinst.ini`.
```ini
[PostgreSQL]
Driver              = PostgreSQL Driver
Database            = test_db
Servername          = localhost
UserName            = postgres
Password            = password
Port                = 5432

[SQLite]
Driver          = SQLite Driver
Database=/tmp/testing
```

See also: [unixODBC without the GUI](http://www.unixodbc.org/odbcinst.html) for more information and examples.

## Usage
All of the following examples assume you have already created a query `con`.
See [Connecting to a database](#connecting-to-a-database) for more information on establishing a
connection.

### Table and Field information
`dbListTables()` is used for listing tables in a database.
```r
dbListTables(con)

# List tables beginning with f
dbListTables(con, table_name = "f%")

# List all fields in the 'flights' database
dbListFields(con, "flights")
```

### Reading
`dbReadTable()` will read a full table into an R `data.frame()`.
```r
data <- dbReadTable(con, "flights")
```

### Writing
`dbWriteTable()` will write an R `data.frame()` to an SQL table.
```r
data <- dbWriteTable(con, "iris", iris)
```

### Querying
`dbGetQuery()` will submit a query and fetch the results. It is also possible
to submit the query and fetch separately with `dbSendQuery()` and `dbFetch()`.
If the `n=` argument to `dbFetch()` can be used to fetch only part of a query.
```r
result <- dbSendQuery(con, "SELECT flight, tailnum, origin FROM flights ORDER BY origin")

# Retrieve the first 100 results
first_100 <- dbFetch(result, n = 100)

# Retrieve the rest of the results
rest <- dbFetch(result)
```

## Benchmarks

Odbconnect is often much faster than the existing
[RODBC](https://cran.r-project.org/package=RODBC) and DBI compatible
[RODBCDBI](https://cran.r-project.org/package=RODBCDBI) packages.

### Reading
Reading a from a PostgreSQL database with the nytflights13 'flights' database (336,776 rows, 19 columns).
```{r, cache = TRUE}
# First using RODBC / RODBCDBI
library(DBI)
library(RODBCDBI)
rodbc <- dbConnect(RODBCDBI::ODBC(), dsn = "PostgreSQL")
system.time(rodbc_result <- dbReadTable(rodbc, "flights"))

# Now using odbconnect
library(odbconnect)
odbconnect <- dbConnect(odbconnect::odbconnect(), dsn = "PostgreSQL")
system.time(odbconnect_result <- dbReadTable(odbconnect, "flights"))

library(tibble)
as_tibble(odbconnect_result)

identical(dim(rodbc_result), dim(odbconnect_result))
rm(rodbc_result, odbconnect_result, odbconnect, rodbc)
gc(verbose = FALSE)
```
### Writing
Writing the same dataset to the database.
```{r, echo = FALSE, cache = TRUE}
odbconnect <- dbConnect(odbconnect::odbconnect(), dsn = "PostgreSQL")
if (dbExistsTable(odbconnect, "flights2")) { dbRemoveTable(odbconnect, "flights2") }
if (dbExistsTable(odbconnect, "flights3")) { dbRemoveTable(odbconnect, "flights3") }
rm(odbconnect)
gc()
```

```{r, cache = TRUE}
library(nycflights13)
# rodbc does not support writing timestamps natively.
rodbc <- dbConnect(RODBCDBI::ODBC(), dsn = "PostgreSQL")
system.time(dbWriteTable(rodbc, "flights2", as.data.frame(flights[, names(flights) != "time_hour"])))

# Now using odbconnect
odbconnect <- dbConnect(odbconnect::odbconnect(), dsn = "PostgreSQL")
system.time(dbWriteTable(odbconnect, "flights3", as.data.frame(flights)))
```
