---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```


# odbconnect
[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)
[![Travis-CI Build Status](https://travis-ci.org/rstats-db/odbconnect.svg?branch=master)](https://travis-ci.org/rstats-db/odbconnect)
[![Coverage Status](https://img.shields.io/codecov/c/github/rstats-db/odbconnect/master.svg)](https://codecov.io/github/rstats-db/odbconnect?branch=master)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/rstats-db/odbconnect?branch=master&svg=true)](https://ci.appveyor.com/project/rstats-db/odbconnect)

The goal of odbconnect is to provide a DBI-compliant interface to ODBC drivers.
The current implementation uses the [nanodbc](http://nanodbc.lexicalunit.com/)
C++ library.

## Installation

ODBC drivers should be compiled against unixodbc on linux and MacOS. Drivers
compiled against iODBC may work, but are not supported.

### MacOS
[homebrew](http://brew.sh/) can be used to easily install database drivers.

```shell
# Install the unixodbc library
brew install unixodbc

# PostgreSQL ODBC ODBC Drivers
brew install psqlodbc

# MySQL ODBC Drivers
brew install mysql-ODBC Drivers-c

# SQLite ODBC Drivers
brew install sqliteodbc

# SQL Server ODBC Drivers (Free TDS)
brew install freetds --with-unixodbc
```

### Linux - Ubuntu / Debian
```shell
# Install the unixodbc library
apt-get install unixodbc unixodbc-dev

# PostgreSQL ODBC ODBC Drivers
apt-get install odbc-postgresql

# MySQL ODBC Drivers
apt-get install libmyodbc

# SQLite ODBC Drivers
apt-get install libsqliteodbc

# SQL Server ODBC Drivers (Free TDS)
apt-get install tdsodbc
```

### R
```R
# install.packages(devtools)
devtools::install_github("rstats-db/odbconnect")
```

## Benchmarks vs RODBC / RODBCDBI

### Reading
Simply reading a postgres table with the nytflights13 'flights' database.
```{r}
# First using RODBC / RODBCDBI
library(DBI)
library(RODBCDBI)
rodbc <- dbConnect(RODBCDBI::ODBC(), dsn = "PostgreSQL")
system.time(rodbc_result <- dbReadTable(rodbc, "flights"))

# Now using odbconnect
library(odbconnect)
odbconnect <- dbConnect(odbconnect::odbconnect(), dsn = "PostgreSQL")
system.time(odbconnect_result <- dbReadTable(odbconnect, "flights"))

library(tibble)
as_tibble(odbconnect_result)

identical(dim(rodbc_result), dim(odbconnect_result))
rm(rodbc_result, odbconnect_result)
gc()
```
### Writing
Writing the flights database works as well.
```{r}
library(nycflights13)
# rodbc does not support writing timestamps natively.
system.time(dbWriteTable(rodbc, "flights2", as.data.frame(flights[, names(flights) != "time_hour"])))

# Now using odbconnect
odbconnect <- dbConnect(odbconnect::odbconnect(), dsn = "PostgreSQL")
system.time(dbWriteTable(odbconnect, "flights3", as.data.frame(flights)))
```
## ODBC Documentation

https://msdn.microsoft.com/en-us/library/ms712628(v=vs.85).aspx
https://msdn.microsoft.com/en-us/library/ms714086(v=vs.85).aspx
