---
title: "Installing and Configuring Drivers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing and Configuring Drivers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette outlines how to install common ODBC drivers and configure them with a driver manager.

## Installation

For Unix and MacOS, ODBC drivers should be compiled against [unixODBC](https://www.unixodbc.org/). Drivers compiled against [iODBC](https://www.iodbc.org/) _may_ also work, but are not fully supported.

After installation of the driver manager and driver, you will have to register the driver in a [odbcinst.ini](#dsn-configuration-files) file for it to appear in `odbc::odbcListDrivers()`.

### Windows

Windows is bundled with ODBC libraries. However, drivers for each database need to be installed separately. Windows ODBC drivers typically include an installer that needs to be run and will install the driver to the proper locations.

### MacOS

[homebrew](https://brew.sh/) can be used to easily install database drivers on MacOS.

The UnixODBC driver manager is required for all databases:

```shell
## Install the unixODBC library
brew install unixodbc
```

Most common database drivers can be installed with homebrew. Some drivers are only available from the database vendor.

```shell
## Microsoft SQL Server ODBC drivers
brew tap microsoft/mssql-release https://github.com/Microsoft/homebrew-mssql-release
brew update
brew install msodbcsql17 mssql-tools

## SQL Server ODBC drivers (Free TDS)
brew install freetds

## PostgreSQL ODBC drivers
brew install psqlodbc

## MySQL ODBC drivers (and database)
brew install mysql

## SQLite ODBC drivers
brew install sqliteodbc
```

Some driver installations will not automatically configure the driver with the driver manager. See the ["Connecting to a Database"](#connecting) section below to learn more about configuring drivers and data sources.

### Linux - Debian / Ubuntu

[apt-get](https://wiki.debian.org/Apt) can be used to easily install database
drivers on Linux distributions which support it, such as Debian and Ubuntu.

The UnixODBC driver manager is required for all databases:

```shell
## Install the unixODBC library
apt-get install unixodbc unixodbc-dev
```

For SQL Server on Linux it is recommended you use the [Microsoft Drivers](https://learn.microsoft.com/en-us/sql/connect/odbc/linux-mac/installing-the-microsoft-odbc-driver-for-sql-server?view=sql-server-ver15#ubuntu18). The code below will install some common drivers:

```shell
## SQL Server ODBC Drivers (Free TDS)
apt-get install tdsodbc

## PostgreSQL ODBC ODBC Driversz
apt-get install odbc-postgresql

## MySQL ODBC Drivers
apt-get install libmyodbc

## SQLite ODBC Drivers
apt-get install libsqliteodbc
```

## Driver config

On Windows, driver config is managed by the operating system, and you don't generally need to edit any configuration files directly. On MacOS and Linux, however, the driver config is managed by unixODBC, and depending on how the driver is installed, it may or may not be automatically configured. 

You can see which drivers are configured by running `odbcListDrivers()`. If the driver you want is not installed, you will need to edit `odbcinst.ini`. You can use `odbcinst -j` (provided by unixODBC) to find out where this file lives.

The basic form of this file is something like this:

```ini
[PostgreSQL Driver]
Driver          = /usr/local/lib/psqlodbcw.so

[SQLite Driver]
Driver          = /usr/local/lib/libsqlite3odbc.dylib
```

i.e., a mapping between driver names, and the location of the driver file. The driver name is what you pass to `odbc::odbc()` when you connect, e,g:

```{r}
#| eval: false

library(DBI)
con1 <- dbConnect(odbc::odbc(), driver = "PostgreSQL Driver")
con2 <- dbConnect(odbc::odbc(), driver = "SQLite Driver")
```

If installing the driver did not automatically update this file, you'll need to update it yourself. You'll need to figure out where the driver library was installed, using something like `brew list {drivername}` or `dpkg-query -L {drivername}`.

## Datasource config

It's also possible to setup named 

#### odbcinst.ini

`odbcinst.ini` contains driver information, particularly the _driver name_ and the location of the installed driver file. Multiple drivers can be specified in the same file. For example:

On MacOS aarch64 machines, drivers installed via homebrew are in a different location. Those might look something like:

```ini
[PostgreSQL Driver]
Driver          = /opt/homebrew/lib/psqlodbcw.so

[SQLite Driver]
Driver          = /opt/homebrew/lib/libsqlite3odbc.dylib
```

### Data sources

It's also possible to configure named data sources, so instead of typing this:

```{r}
#| eval: false

library(DBI)
con <- dbConnect(
  odbc::odbc(),
  driver = "PostgreSQL Driver",
  database = "test_db",
  username = "postgres",
  password = Sys.getenv("PASSWORD"),
  host = "localhost",
  port = 5432
)
```

You can type:

```{r}
#| eval: false

con <- dbConnect(odbc::odbc(), dsn = "test_db")
```

On Windows, you can set up a data source by using the [ODBC Data Source Administrator](https://learn.microsoft.com/en-us/sql/odbc/admin/odbc-data-source-administrator) application is used to manage ODBC data sources on Windows. On Linux and Mac, you'll need to edit `odbc.ini` (again finding the location with `odbcinst -j`). Each data source is defined in a separate section (defined by `[name]`]), with each paramter going on its own line, e.g.:

```ini
[test_db]
driver     = PostgreSQL Driver
database   = test_db
username   = postgres
password   = password
host       = localhost
port       = 5432
```

You can see all currently defined data sources by running `odbcListDataSources()`.
